---
title: "Important functions for data wrangling"
author: "Template: Ian Hussey; Content: [Tim Schneeberger]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

# set knit options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

# Instructions

Most data wrangling tasks can be accomplished with a relatively small number of functions. I've listed the most important ones here. 

Use this file to keep notes about what a given function does in your own words, the situations where you would use it, and working examples. You can make use of built-in datasets to do this (e.g., `mtcars`) or load external data sets (although its easier to break the examples if you modify the data etc.).

Learning how to wrangle data efficiently involves a combination of knowing how to break the problem down into smaller components; knowing which functions are available to accomplish each component operation and how to use them; how to search documentation to learn or refresh your knowledge of how a function works; and how to discover new functions that accomplish new component operations. 

Remember that you can look up the help documentation for any function by typing a question mark followed by its name in the console, e.g., `?dplyr::mutate`. The help documentation provides details of a function's parameters and defaults, its outputs, and examples of its use. Note that when you can also open the help documentation for an entire package by typing the package name, e.g., `?dplyr`. This can be very useful to discover what other functions that package has: scroll down to the bottom of any help page and click the "Index" link to see all help pages for that package.

I have prepended each function below with package it comes from so that you know. For example, `summarize` is listed as `dplyr::summarize`. Usually you don't have to do this when using a function, although you can use this to resolve a common bug known as name conflicts (see [this blog post](https://www.r-bloggers.com/2010/08/namespaces-and-name-conflicts/) for discussion). 

# Resources (for this session and others)

- You can find cheatsheets for the dplyr, tidyr, and RMarkdown in the /resources/cheatsheets folder.
- The Open Source textbook R for Data Science (aka, [Wickham's R4DS](https://r4ds.hadley.nz/)) is invaluable. Hadley Wickham is the main developer of the "tidyverse" set of packages, including dplyr, tidyr, ggplot2, stringr, lubridate, and others. See its [section on data transformation](https://r4ds.hadley.nz/data-transform). 
  - The entire second edition of the book is available at [https://r4ds.hadley.nz/](https://r4ds.hadley.nz/).
  - The first edition is also available. It does some things better in my opinion, e.g., it has a better explanation of the pipe (`%>%` or `|>`). See [https://r4ds.had.co.nz/pipes.html](https://r4ds.had.co.nz/pipes.html). 
  - The first edition also talks about RMarkdown, whereas the second edition has moved to a different technology called Quarto (which we won't cover, although they're similar). See [https://r4ds.had.co.nz/r-markdown.html](https://r4ds.had.co.nz/r-markdown.html).
- For those of you who prefer to learn in an interactive environment, I now suggest this web app over Swirl as it is more user-friendly: [https://allisonhorst.shinyapps.io/dplyr-learnr/#section-welcome](https://allisonhorst.shinyapps.io/dplyr-learnr/#section-welcome).
- For those of you who prefer some video content - although seeing other people code can never replace practicing coding yourself! - I can also recommend De Bruine et al.'s Open Source textbook and videos [Data Skills for Reproducible Research](https://psyteachr.github.io/reprores-v3/). E.g., see their page with links to videos for [dplyr](https://psyteachr.github.io/reprores-v3/dplyr.html) and [tidyr](https://psyteachr.github.io/reprores-v3/tidyr.html). 

# Dependencies

The packages these functions come from, which must be loaded to use them.

```{r}

library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(janitor)

```

# Basics

## read_csv 

vs. `read.csv()`

*Always, always, always* use relative paths rather than absolute paths.

- Absolute path (bad): "~/Ian/Desktop/my_project/data/data_raw.csv"
- Relative path (good): "../data/data_raw.csv"

When using relative paths: "../" means up one directory level, whereas "directory_name/" goes down one directory level.

Relative paths will work on other machines, where absolute paths will break. 

Relative paths only work in .Rmd files and not .R files. Even without RMarkdown's other benefits, this makes them worth using.

Useful arguments:

- `skip` can be used to skip the first N rows while reading the data

```{r}



# readr::read_csv()

```

## dir.create

```{r}

# dir.create()

```

## colnames & dput

```{r}

# colnames()

# dput(colnames())

```

# Wrangling

## the pipe: `%>%` or `|>`

```{r}

# %>%

# |>

```

## round_half_up

```{r}

# janitor::round_half_up()

```

## clean_names

```{r}

# janitor::clean_names()

```

## filter

```{r}

# dplyr::filter()
## Multiple criteria, 'and' or 'or' combinations

#You can also have multiple criteria in your filter call, both of which have to be met (x `&` y), or either one of which have to be met (x `|` y). Bsp: 
example_multiple_criteria_1 <- data_amp_trimmed %>%
  filter(block != "test" & correct == 1)

example_multiple_criteria_2 <- data_amp_trimmed %>%
  filter(block != "test" | correct == 1)

# note that these provide different results - make sure you understand why
identical(example_multiple_criteria_1, example_multiple_criteria_2)

```

## slice

Slice is for retaining or dropping rows from a df.

We might use it just after reading in files if there are junk rows at the header.

```{r}

# dplyr::slice()

mtcars # this should have 32 rows
slice(mtcars, 2:n()) # this should have 31 rows

```

## select

```{r}

# dplyr::select()

```

## rename

```{r}

# dplyr::rename()

```

## mutate

```{r}

# dplyr::mutate()
# clean up the gender variable
data_demographics_gender_tidy_2 <- data_demographics_trimmed %>%
  # filter only the gender item, not age
  filter(item == "gender") %>%
  # change the name of the response variable to what it now represents: gender
  rename(gender = response) %>%
  # change or remove weird responses to the gender question
  mutate(gender = str_to_lower(gender),
         gender = str_remove_all(gender, "[\\d.]"), # remove everything except letters
         gender = na_if(gender, ""), 
         gender = case_when(gender == "woman" ~ "female",
                            gender == "man" ~ "male",
                            gender == "girl" ~ "female",
                            gender == "yes" ~ NA_character_,
                            gender == "dude" ~ "male",
                            gender == "non binary" ~ "non-binary",
                            TRUE ~ gender)) %>%
```

## case_when

```{r}

# dplyr::case_when() # instead of ifelse()
#`case_when()` allows you to compare multiple logical tests or if-else tests.

The AMP data needs to be reverse scored. Just like an item on a self-report that is worded negatively (e.g., most items: I am a good person; some items: I am a bad person), the negative prime trials have the opposite 'accuracy' values that they should. Use `mutate()` and `case_when()` to reverse score the negative prime trials, so that what was 0 is now 1 and what was 1 is now 0.

```{r}

data_amp_tidy <- data_amp_test_trials_with_fast_trials %>%
  mutate(correct = case_when(trial_type == "prime_positive" ~ correct,
                             trial_type == "prime_negative" & correct == 0 ~ 1,
                             trial_type == "prime_negative" & correct == 1 ~ 0))

# you can also specify a default value to return if none of the logical tests are passed with 'TRUE ~':
data_amp_tidy <- data_amp_test_trials_with_fast_trials %>%
  mutate(correct = case_when(trial_type == "prime_negative" & correct == 0 ~ 1,
                             trial_type == "prime_negative" & correct == 1 ~ 0,
                             TRUE ~ correct))
```

## summarize

```{r}

# dplyr::summarize() 
# Summarizing across rows

It is very common that we need to create summaries across rows. For example, to create the mean and standard deviation of a column like age. This can be done with `summarize()`. Remember: `mutate()` creates new columns or modifies the contents of existing columns, but does not change the number of rows. Whereas `summarize()` reduces a data frame down to one row.

```{r}

# mean
data_demographics_age_tidy %>%
  summarize(mean_age = mean(age, na.rm = TRUE))

# SD
data_demographics_age_tidy %>%
  summarize(sd_age = sd(age, na.rm = TRUE))

# mean and SD with rounding, illustrating how multiple summarizes can be done in one function call
data_demographics_age_tidy %>%
  summarize(mean_age = mean(age, na.rm = TRUE),
            mean_age = round_half_up(mean_age, digits = 2),
            sd_age = sd(age, na.rm = TRUE),
            sd_age = round_half_up(sd_age, digits = 2))
#gives you an output of the mean and the sd
```

## count

```{r}

# dplyr::count()

```

## distinct

```{r}

# dplyr::distinct()

```

## group_by

```{r}

# dplyr::group_by() # or ".by =" 

```

## rowwise

```{r}

# dplyr::rowwise()

```

## lead

```{r}

# dplyr::lead()

```

## lag

```{r}

# dplyr::lag()

```

## joins

```{r}

# dplyr::full_join()

# dplyr::left_join()

# dplyr::right_join()

# dplyr::inner_join()

# dplyr::semi_join()

```

## pivots

```{r}

# tidyr::pivot_wider()

# tidyr::pivot_longer()

```

## drop_na

```{r}

# tidyr::drop_na()

```

## separate

```{r}

# tidyr::separate()

```

## fill

```{r}

# tidyr::fill()

```

# Printing tables

```{r}

# mtcars |> # example data
#   knitr::kable() |> # print a nicer looking table
#   kableExtra::kable_classic(full_width = FALSE) # print nicer again

```

# Other packages 

Other packages you may need for wrangling which aren't covered here:

- library(forcats). Everything to do with factors and factor levels. Useful for plotting and establishing reference levels for statistical tests.
- library(stringr). Everything to do with strings, searching for strings, modifying strings, parsing them, etc.
- library(lubridate). Everything to do with dates, parsing dates, converting date formats, etc. 

# Session info

```{r}

sessionInfo()

```



